\chapter{Bases of a Vector Space (Part B)}
\section{The Dimension Theorem}

\subsection{The Strong Exchange Lemma}

Before we establish the main theorem of this chapter, we formalize a stronger version of the Steinitz Exchange Theorem (Lemma D from the previous section). We proved that an independent list cannot be larger than a spanning list. However, the exact mechanism of the proof actually guarantees something even more powerful: we can precisely substitute the independent vectors into the spanning set.

\begin{lemma}[Lemma D': Strong Steinitz Exchange]
Let $V$ be a vector space over $K$, and suppose that $V = \Sp(v_1, \dots, v_n)$ for some list of vectors $v_1, \dots, v_n \in V$. Let $u_1, \dots, u_m \in V$ be a linearly independent list. 

Then $m \le n$, and moreover, one can delete exactly $m$ vectors from the spanning list $v_1, \dots, v_n$ such that the remaining $n-m$ vectors, together with all the vectors $u_1, \dots, u_m$, still span $V$.
\[
\Sp(u_1, \dots, u_m, v'_1, \dots, v'_{n-m}) = V
\]
(where the $v'$ vectors are the survivors of the original spanning list).
\end{lemma}

\begin{proof}
The fact that $m \le n$ was already proven in Lemma D. To prove the "moreover" part, we simply apply the exact step-by-step substitution procedure described in the proof of Lemma D. 
For steps $j = 1, 2, \dots, m$, we sequentially add $u_j$ to the front of the list. At each step, the list becomes dependent (Lemma C), and we forcefully eject a redundant $v_k$ vector (Lemma B) to restore independence while perfectly preserving the span. After $m$ steps, we have injected all $m$ independent vectors and deleted exactly $m$ of the original spanning vectors. The resulting mixed list still spans $V$.
\end{proof}

\subsection{The Existence and Uniqueness of Dimension}

We are now ready to state and prove the foundational theorem of finite-dimensional vector spaces. It guarantees that bases always exist and that they provide an absolute, rigid measure of the "size" of a vector space.

\begin{theorem}[The Basis Theorem]
Let $V$ be a finite-dimensional vector space over $K$. Then $V$ has a basis with finitely many elements. 

Moreover, \textbf{every} basis of $V$ is finite, and any two bases of $V$ have exactly the same number of elements.
\end{theorem}

\begin{remark}[Infinite Dimensional Spaces]
This theorem is actually true even if $V$ is not finite-dimensional (this is known as the "Hamel Basis Theorem" and requires Zorn's Lemma). However, in this course, we will only prove the theorem for finite-dimensional spaces. The infinite-dimensional case is rigorously covered in advanced algebra courses (e.g., \textit{Grundstrukturen} in the 2nd semester).
\end{remark}

\subsection{Proof of the Basis Theorem}

Because this theorem is so monumental, we will break its proof down into three distinct, manageable statements. 

\subsubsection*{Statement 1: V has a finite basis.}

\begin{proof}
Because we assumed $V$ is finite-dimensional, there must exist some finite list of vectors that spans $V$. Let us call this spanning set $S = \{w_1, \dots, w_k\}$. If $S$ is already linearly independent, it is a basis, and we are done.

If $S$ is linearly dependent, we can extract a basis from it using the following systematic reduction algorithm (often called a "Redundancy Filter"):

\begin{center}
\resizebox{0.9\textwidth}{!}{
    \begin{tikzpicture}[scale=1,
        node distance=1.5cm,
        box/.style={draw, thick, rounded corners, align=center, inner sep=2ex},
        arrow/.style={->, thick, >=stealth}
    ]
        \node[box, fill=blue!10] (start) {\textbf{Spanning List} \\ $w_1, w_2, \dots, w_k$};
        
        \node[box, fill=yellow!10, below=1cm of start] (check) {Read left to right. \\ Is $w_j \in \Sp(w_1, \dots, w_{j-1})$?};
        
        \node[box, fill=red!10, left=1cm of check, yshift=-1.5cm] (drop) {\textbf{Yes:} Drop $w_j$. \\ (Span is preserved by Lemma B)};
        \node[box, fill=green!10, right=1cm of check, yshift=-1.5cm] (keep) {\textbf{No:} Keep $w_j$.};
        
        \node[box, fill=blue!10, below=4cm of start] (final) {\textbf{Final Output} \\ A linearly independent list \\ that still spans $V$ (A Basis!)};
        
        \draw[arrow] (start) -- (check);
        \draw[arrow] (check) -- node[above left] {Redundant} (drop);
        \draw[arrow] (check) -- node[above right] {Independent} (keep);
        \draw[arrow] (drop) |- (final);
        \draw[arrow] (keep) |- (final);
    \end{tikzpicture}
}
\end{center}

Specifically:
1. If the zero vector $0_V$ is in the list, drop it.
2. Read the list from left to right. If a vector $w_j$ is a linear combination of its strictly preceding vectors (i.e., $w_j \in \Sp(w_1, \dots, w_{j-1})$), drop it. 
By Lemma B, dropping this redundant vector does not change the span of the list. We repeat this until we have checked every vector. The remaining list still spans $V$. Furthermore, by Proposition 9.1, it must be linearly independent, because we systematically deleted every vector that could be written as a combination of the others. Thus, the remaining list is a finite basis.
\end{proof}

\subsubsection*{Statement 2: Every basis of V is finite.}

\begin{proof}
We prove this by contradiction. Suppose there exists a basis $\mathcal{A}$ of $V$ that is \textit{not} a finite set. 

From Statement 1, we know that $V$ definitely possesses at least one finite basis. Let us call this finite basis $S = \{v_1, \dots, v_n\}$. The size of this finite basis is exactly $n$.

Because $\mathcal{A}$ is infinite, it contains infinitely many linearly independent vectors. Therefore, we can arbitrarily select $n + 100$ distinct vectors from $\mathcal{A}$. Let us call them $u_1, \dots, u_{n+100}$. 
Because they belong to a basis, this chosen list of $n+100$ vectors is strictly linearly independent. 

We now have two lists:
\begin{itemize}
    \item A spanning list $S$ containing exactly $n$ elements.
    \item An independent list containing $n+100$ elements.
\end{itemize}

By the Steinitz Exchange Theorem (Lemma D), the number of elements in any independent list must be less than or equal to the number of elements in any spanning list. Therefore, we must have:
\[
n + 100 \le n
\]
This implies $100 \le 0$, which is a blatant mathematical contradiction! Thus, our initial assumption must be wrong. Every basis of $V$ must be finite.
\end{proof}

\subsubsection*{Statement 3: Any two bases have the same cardinality.}

\begin{proof}
Let $\mathcal{A}$ and $\mathcal{B}$ be two arbitrary bases of $V$. By Statement 2, we know that both $\mathcal{A}$ and $\mathcal{B}$ must be finite sets. Let $|\mathcal{A}|$ denote the number of elements in $\mathcal{A}$, and $|\mathcal{B}|$ denote the number of elements in $\mathcal{B}$.

Because they are bases, both sets are simultaneously independent and spanning. This allows us to apply Lemma D twice in alternating directions:

\begin{center}
\begin{tikzpicture}[
    node distance=1.5cm,
    box/.style={draw, thick, rounded corners, align=center, inner sep=2ex},
    arrow/.style={->, thick, >=stealth}
]
    \node[box, fill=gray!10] (A) {\textbf{Basis $\mathcal{A}$}};
    \node[box, fill=gray!10, right=4cm of A] (B) {\textbf{Basis $\mathcal{B}$}};
    
    \draw[arrow] (A.north east) to[bend left=20] node[midway, above] {Independent in Spanning} node[midway, below] {$|\mathcal{A}| \le |\mathcal{B}|$} (B.north west);
    
    \draw[arrow] (B.south west) to[bend left=20] node[midway, below] {Independent in Spanning} node[midway, above] {$|\mathcal{B}| \le |\mathcal{A}|$} (A.south east);
\end{tikzpicture}
\end{center}

\begin{enumerate}
    \item Because $\mathcal{A}$ is linearly independent and $\mathcal{B}$ spans $V$, Lemma D dictates that $|\mathcal{A}| \le |\mathcal{B}|$.
    \item Reversing their roles, because $\mathcal{B}$ is linearly independent and $\mathcal{A}$ spans $V$, Lemma D dictates that $|\mathcal{B}| \le |\mathcal{A}|$.
\end{enumerate}

Since $|\mathcal{A}| \le |\mathcal{B}|$ and $|\mathcal{B}| \le |\mathcal{A}|$, it must be that $|\mathcal{A}| = |\mathcal{B}|$. All bases contain exactly the same number of elements.
\end{proof}

\subsection{The Concept of Dimension}

Because the number of elements in a basis is a rigid, absolute invariant of the vector space, we can finally define the space's dimension.

\begin{definition}[Dimension]
Let $V$ be a finite-dimensional vector space over $K$. We define the \textbf{dimension} of $V$ to be the number of elements in any basis of $V$. 
We denote this integer by $\dim V = n \in \mathbb{Z}_{\ge 0}$.

If $V$ is not finite-dimensional, we write $\dim V = \infty$.
\end{definition}

\begin{remark}[The Zero Space]
Consider the zero vector space $V = \{0_V\}$. Its only spanning set is $\{0_V\}$, which is linearly dependent. The only linearly independent subset that spans it is the empty set $\emptyset$. 
Therefore, the basis of the zero space is $\emptyset$. Because the empty set contains $0$ elements, we have:
\[
\dim \{0_V\} = 0
\]
\end{remark}

\subsection{Examples of Dimensions}

Now that we have formally defined dimension, we can readily determine the dimensions of our familiar vector spaces by simply counting the elements in their standard bases.

\begin{example}[The Coordinate Space]
The standard basis for $K^n$ is $\{e_1, \dots, e_n\}$. Because this set contains exactly $n$ elements, we have:
\[
\dim K^n = n
\]
\end{example}

\begin{example}[Bounded Polynomials]
The standard basis for the space of polynomials of degree $\le d$ is $\{1, x, x^2, \dots, x^d\}$. Be careful when counting here! Because we start at $x^0 = 1$, this set contains $d+1$ elements. Thus:
\[
\dim K[x]_d = d + 1
\]
\end{example}

\begin{example}[Matrices]
The standard basis for the matrix space $\M_{m \times n}(K)$ consists of the individual single-entry matrices $E_{i,j}$. Since there are $m$ rows and $n$ columns, there are exactly $m \cdot n$ such matrices. Thus:
\[
\dim \M_{m \times n}(K) = m \cdot n
\]
\end{example}

\begin{example}[Function Spaces]
Let $S$ be an arbitrary set. Recall that the space of all functions from $S$ to $K$, denoted as $K^S := \{f \mid f: S \to K\}$, forms a vector space over $K$. 

The dimension of this space matches the cardinality of the underlying set $S$:
\[
\dim K^S = |S|
\]

\begin{exercise}
If $S$ is a finite set, say $S = \{s_1, \dots, s_n\}$, can you construct an explicit basis for $K^S$? \\
\textit{(Hint: Try to define a family of functions $f_i$ where each function outputs $1$ for exactly one specific element $s_i$, and $0$ for all other elements).}
\end{exercise}

\begin{remark}
When the domain $S$ is an infinite set, the space $K^S$ is infinite-dimensional. In this case, we unfortunately do not have a "nice" or easily describable basis for it.
\end{remark}
\end{example}