\setcounter{chapter}{5}
\renewcommand{\thetheorem}{\thechapter.\arabic{theorem}}
\setcounter{theorem}{0}

\chapter{Vector Spaces}

\subsection{Definition and Axioms}

\begin{definition}[Vector Space]
A \textit{vector space} over a field $K$ is a set $V$, endowed with two operations:
\begin{align*}
+: V \times V &\longrightarrow V, \quad (v_1, v_2) \mapsto v_1 + v_2  &\text{(Addition of vectors)} \\
\cdot: K \times V &\longrightarrow V, \quad (\alpha, v) \mapsto \alpha \cdot v  &\text{(Scalar multiplication)}
\end{align*}
such that the following conditions (axioms) hold:
\begin{description}
    \item[\textbf{V1}] $\forall v_1, v_2, v_3 \in V$: $v_1 + (v_2 + v_3) = (v_1 + v_2) + v_3$.
    \item[\textbf{V2}] $\exists$ an element $0 \in V$ such that $0 + v = v + 0 = v$ for all $v \in V$. (Sometimes denoted $0_V$).
    \item[\textbf{V3}] $\forall v \in V, \exists v' \in V$ such that $v + v' = 0$.
    \item[\textbf{V4}] $\forall v_1, v_2 \in V$: $v_1 + v_2 = v_2 + v_1$.
    \item[\textbf{V5}] $\forall a, b \in K, \forall v \in V$: $a \cdot (b \cdot v) = (a \cdot b) \cdot v$.
    \item[\textbf{V6}] $\forall v \in V$: $1 \cdot v = v$.
    \item[\textbf{V7}] $\forall \alpha \in K, \forall v_1, v_2 \in V$: $\alpha \cdot (v_1 + v_2) = \alpha \cdot v_1 + \alpha \cdot v_2$.
    \item[\textbf{V8}] $\forall \alpha_1, \alpha_2 \in K, \forall v \in V$: $(\alpha_1 + \alpha_2) \cdot v = \alpha_1 \cdot v + \alpha_2 \cdot v$.
\end{description}
\end{definition}

\begin{lemma}
Let $V$ be a vector space over a field $K$. Then the following properties hold:
\begin{enumerate}
    \item The $0$ vector from (V2) is unique. To avoid confusion (e.g., with $0 \in K$), we will sometimes denote it by $0_V$.
    \item The element $v'$ (corresponding to $v$) from (V3) is unique. We will denote it by $-v$. We will also write $v - w := v + (-w)$ for all $v, w \in V$.
    \item For all $v \in V$, we have $0 \cdot v = 0_V$ (i.e., $0_K \cdot v = 0_V$).
    \item For all $\alpha \in K$, we have $\alpha \cdot 0_V = 0_V$.
    \item $(-1) \cdot v = -v$ for all $v \in V$.
\end{enumerate}
\end{lemma}

\begin{proof}
\textbf{(a)} Assume there are two such zero elements, $0_1$ and $0_2$. 
Then, by applying (V2) to both, we get: $0_1 = 0_1 + 0_2 = 0_2 + 0_1 = 0_2$. Thus, $0_1 = 0_2$.

\textbf{(b)} Assume there are two inverses $v_1'$ and $v_2'$ for an element $v$. 
Then we can write: 
\[
v_1' = v_1' + 0_V = v_1' + (v + v_2') = (v_1' + v) + v_2' = 0_V + v_2' = v_2'
\]
Thus, the inverse is unique.

\textbf{(c)} Consider $0 \cdot v = (0 + 0) \cdot v = 0 \cdot v + 0 \cdot v$. 
By adding $-(0 \cdot v)$ to both sides, we yield $0_V = 0 \cdot v$.

\textbf{(d)} Similarly, $\alpha \cdot 0_V = \alpha \cdot (0_V + 0_V) = \alpha \cdot 0_V + \alpha \cdot 0_V$. Adding $-(\alpha \cdot 0_V)$ yields $0_V = \alpha \cdot 0_V$.

\textbf{(e)} Observe that $v + (-1) \cdot v = 1 \cdot v + (-1) \cdot v = (1 - 1) \cdot v = 0 \cdot v = 0_V$. Since the inverse is unique by part (b), it must be that $(-1) \cdot v = -v$.
\end{proof}

\subsection{Examples of Vector Spaces}

\subsubsection{The Space $K^n$}
\begin{example}
The set of $n$-tuples $K^n$ becomes a vector space over $K$ if we endow it with the following operations. 
Let $v = (a_1, \dots, a_n)$ and $w = (b_1, \dots, b_n) \in K^n$. We define addition in $K^n$ component-wise:
\[
v + w = (a_1, \dots, a_n) + (b_1, \dots, b_n) := (a_1 + b_1, \dots, a_n + b_n)
\]
For $\alpha \in K$, we define scalar multiplication as:
\[
\alpha \cdot v = \alpha \cdot (a_1, \dots, a_n) := (\alpha \cdot a_1, \dots, \alpha \cdot a_n)
\]
The zero vector is explicitly defined as $0 := (0, \dots, 0) \in K^n$. 
(The proof that this constitutes a vector space relies directly on the fact that $K$ itself satisfies the axioms of a field).
\end{example}

\subsubsection{Spaces of Matrices}
\begin{example}
Let $M_{m \times n}(K)$ be the set of all $m \times n$ matrices with entries in $K$. We define addition of matrices component-wise: if $A = (A_{i,j})$ and $B = (B_{i,j})$, then $A + B = (A_{i,j} + B_{i,j})$. Similarly, scalar multiplication is defined component-wise: $\alpha \cdot A = (\alpha \cdot A_{i,j})$. With these operations, $M_{m \times n}(K)$ forms a vector space over $K$. The zero vector is the $m \times n$ matrix consisting entirely of zeros.
\end{example}

\subsubsection{Function Spaces $K^S$}
\begin{example}
Let $S$ be an arbitrary set. We define the function space:
\[
K^S := \{f: S \to K \mid f \text{ is a function}\}
\]
We define addition $+$ and scalar multiplication $\cdot$ on $K^S$ as follows: 
Let $f, g \in K^S$ and $a \in K$. 
We define $(f + g): S \to K$ point-wise by $x \mapsto f(x) + g(x)$.
We define $(a \cdot f): S \to K$ point-wise by $x \mapsto a \cdot f(x)$.
\end{example}

\subsubsection{Polynomials $K[x]$}
\begin{example}
A polynomial over $K$ is an expression of the form:
\[
f(x) = a_0 + a_1 x + a_2 x^2 + \dots + a_n x^n
\]
where $a_i \in K$. Two polynomials are considered equal if and only if all their corresponding coefficients are equal. Addition and scalar multiplication are performed exactly as expected by manipulating coefficients.

\begin{importantremark}
For finite fields, different polynomials may yield the exact same function! For example, if $K = \mathbb{F}_2$, the polynomials $f(x) = x$ and $g(x) = x^2$ produce the same output values for all inputs in $\mathbb{F}_2$. However, \textbf{as polynomials}, $f(x)$ and $g(x)$ are strictly different mathematical objects!
\end{importantremark}

\textbf{Degree of a polynomial:}
Let $f(x) = a_0 + a_1 x + \dots + a_n x^n$ be a polynomial with $a_n \ne 0$. We say that the \textit{degree} of $f$ is $n$, and we write $\deg(f) = n \in \mathbb{Z}_{\ge 0}$.
Alternatively, if $f(x) \ne 0$:
\[
\deg(f) := \max \{k \in \mathbb{Z}_{\ge 0} \mid a_k \ne 0\}
\]
If $a_0 = a_1 = \dots = a_n = 0$, meaning $f$ is the zero polynomial, we set $\deg(f) = -\infty$. (And consequently, $\deg(f) = -\infty \iff f = 0$).

\begin{remark}
Every polynomial has a well-defined degree.
\end{remark}
\end{example}

\subsection{Linear Subspaces}

\begin{definition}[Linear Subspace]
Let $V$ be a vector space over $K$. A subset $U \subseteq V$ is called a \textit{linear subspace} if it satisfies:
\begin{enumerate}
    \item $0_V \in U$.
    \item $\forall u_1, u_2 \in U \implies u_1 + u_2 \in U$ (closed under addition).
    \item $\forall \alpha \in K, \forall u \in U \implies \alpha \cdot u \in U$ (closed under scalar multiplication).
\end{enumerate}
\end{definition}

\begin{example}
Consider the set $U_b \subseteq K^3$ defined by $U_b = \{(x_1, x_2, x_3) \in K^3 \mid x_1 + x_2 + x_3 = b\}$. 

\textbf{Claim:} $U_b$ is a linear subspace of $K^3$ if and only if $b = 0$.

\begin{proof}
Suppose $U_b$ is a valid subspace. Let $(x_1, x_2, x_3)$ and $(y_1, y_2, y_3) \in U_b$. 
Then by definition, $x_1 + x_2 + x_3 = b$ and $y_1 + y_2 + y_3 = b$. 
Since $U_b$ is a subspace, it must be closed under addition, meaning $(x_1 + y_1, x_2 + y_2, x_3 + y_3) \in U_b$. 
This implies:
\[
(x_1 + y_1) + (x_2 + y_2) + (x_3 + y_3) = b
\]
But rearranging the terms, we also know:
\[
(x_1 + x_2 + x_3) + (y_1 + y_2 + y_3) = b + b = 2b
\]
Therefore, $2b = b$, which implies $b = 0$. 
(Conversely, if $b=0$, it is trivial to check that all three subspace conditions hold).
\end{proof}
\end{example}

\begin{example}[Subspaces of Polynomials]
Fix $d \in \mathbb{Z}_{\ge 0} \cup \{-\infty\}$. 
We define the set:
\[
K[x]_d := \{f \in K[x] \mid \deg(f) \le d\}
\]
(where we agree that $-\infty \le k$ for all $k \in \mathbb{Z}_{\ge 0}$). 

\begin{claim*}
$K[x]_d \subseteq K[x]$ is a linear subspace of $K[x]$.
\end{claim*}

Later, we will see that $K[x]_d$ is a subspace of dimension $d+1$.
\end{example}